---
title: "COVID-19 Case and Vaccination analysis"
date: "8/13/2021"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
library(tidyverse)
library(ggplot2)
library(rpart)
```


# Statement of interest

With this project I want to look into how vaccination rates in the US have 
impacted the last wave (Summer 2021) of the Coronavirus pandemic. 

I want to see if there is a direct relationship between the number of people 
vaccinated and the number of new cases during this wave. 


# Import the data

For my analysis I will use the COVID-19 Data Repository provided by the Johns 
Hopkins University. 
The repository provides data about the cases and deaths and also information 
about the vaccinations. For this project I am interested in US numbers only.


```{r Import, echo=TRUE}
urlpath <- str_c("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/",
"master/csse_covid_19_data/csse_covid_19_time_series/")
url1 <-  str_c(urlpath, "time_series_covid19_confirmed_US.csv")
url2 <-  str_c(urlpath, "time_series_covid19_deaths_US.csv")

vurlpath <- str_c("https://raw.githubusercontent.com/govex/COVID-19/",
                  "master/data_tables/vaccine_data/us_data/time_series/")
vurl1 <- str_c(vurlpath, "people_vaccinated_us_timeline.csv")

covid_cases <- read.csv(url1)
covid_deaths <- read.csv(url2)
head(covid_cases)
head(covid_deaths)

vacc <- read.csv(vurl1)
head(vacc)

```


# Clean up data

The cases and deaths are listed in their respective data frames as columns 
rather than rows. I will pivot the frames to move the relevant data to 
rows, keeping only the columns I am interested in.
Also, I will extract the date in a proper date format.
Then I will combine the cases and deaths together into one data frame,
matching by date and location.

```{r Clean, echo=TRUE}
covid_cases <- covid_cases %>%
    pivot_longer(cols = -c(UID : Combined_Key), 
    names_to = "date", values_to = "cases" ) %>%
    select ( -c(UID,iso2,iso3,code3,FIPS, Lat, Long_, Combined_Key))%>%
    mutate( date = mdy(str_replace(date,'X','')))

covid_deaths <- covid_deaths %>%
    pivot_longer(cols = -c(UID:Population), 
    names_to = "date", values_to = "deaths" ) %>%
    select ( -c(UID,iso2,iso3,code3,FIPS, Lat, Long_, Combined_Key)) %>%
    mutate( date = mdy(str_replace(date,'X','')))

covid_cases <- covid_cases %>%
    left_join(covid_deaths,by = 
                  c("Admin2", "Province_State", "Country_Region", "date"))

```

The vaccination data is already in rows.
I will to keep only the columns I need and transform the date value  and rename
the long column names to make them easier to handle.


```{r}
vacc <- vacc %>%
    select ( -c (FIPS,Lat, Long_, Combined_Key)) %>%
    rename(date = "Date", full = "People_Fully_Vaccinated",
          part = "People_Partially_Vaccinated" ) %>%
    mutate( date = ymd(date)) 
    
```

## Combine data

Unfortunately, the vaccination data is not available per county. 
This means I have to limit my analysis to vaccinations per state.
I will aggregate the cases and deaths per state and date and then merge them
with the vaccinations. 

To account for population differences of the states, I will calculate the 
cases and deaths per 100,000 residents and the vaccinations as a percent of 
the population.

```{r}
state_cases <- covid_cases %>%
    group_by(Province_State, Country_Region, date) %>%
    summarize(cases = sum(cases, na.rm=TRUE), deaths = sum(deaths, na.rm=TRUE),
              Population = sum(Population)) 

state_cases <- state_cases %>%
    left_join(vacc,by = c("Province_State", "Country_Region", "date")) %>%
    mutate(casesph = cases * 100000 / Population,
           deathsph = deaths * 100000 / Population,
           fullp = full * 100 / Population, 
           partp  = part * 100 / Population)
```

# Visualize data to determine target time frame

First I want to look at how the rate of cases and vaccinations changes over time.
This will help me determine the starting point of the summer wave and what time 
frame I want to focus my further analysis on.

I will plot the cases and deaths for the entire US and also look at how the 
vaccination rate changes over time.

```{r}
us_total <- state_cases %>%
    group_by(Country_Region, date) %>%
    summarize(cases = sum(cases), deaths = sum(deaths), 
              Population = sum(Population),
              full = sum(full, na.rm=TRUE), 
              part = sum(full, na.rm=TRUE)) %>%
    mutate(casesph = cases * 100000 / Population,
           deathsph = deaths * 100000 / Population,
           fullp = full * 100 / Population, 
           partp  = part * 100 / Population)
    
tail(us_total)
summary(us_total)
```



Now I will plot the number of cases and death and also draw a plot with the change in
vaccination rates. 

I will set a starting point of 4/1/2021 the date 
vaccinations were widely available in most states.

```{r}
us_total %>%
    filter(date > ymd("2021/4/1")) %>%
    ggplot( aes(x=date, y=casesph)) +
    geom_point(aes(color="Cases")) +
    geom_line(aes(color="Cases")) +
    #scale_y_log10() +
    labs(title = "US Covid-Cases and deaths per 100,000 people", 
         y="Number of cases/deaths", x=NULL)

us_total %>%
    filter(date > ymd("2021/4/1")) %>%
    ggplot( aes(x=date, y=deathsph)) +
    geom_point(aes(color="Deaths")) +
    geom_line(aes(color="Deaths")) +
    #scale_y_log10() +
    labs(title = "US Covid-Cases and deaths per 100,000 people", 
         y="Number of cases/deaths", x=NULL)

us_total %>%
    filter(date > ymd("2021/4/1")) %>%
    ggplot( aes(x=date, y=fullp)) +
    geom_point(aes(y=fullp, color="Fully")) +
    geom_line(aes(y=fullp, color="Fully")) +
    geom_point(aes(y=partp, color="Partially")) +
    geom_line(aes(color="Partially")) +
    #scale_y_log10() +
    labs(title = "Percent of US population Vaccinated", 
         y="% vaccinated", x=NULL)



```

# Analyse differences in states with high and low vaccination rates

The plots show that after leveling off in June, the number of cases 
starts to increase more rapidly again in July. 

I will focus my analysis on the time frame starting on July 1st. The cases as of 
this date will serve as the baseline. 
I will then look at how many cases where added after that date.


## Identify states with low and high vaccination rates

To compare the number of cases in low vs. high vaccinated states, I will select 
the 10 states that had the highest and the lowest vaccination percentage 
on July 1st.

I will then track how many new cases are reported in these states in the 
following weeks.  


```{r}
bottom_10 <- state_cases %>%
    filter(date==ymd("2021/7/1"), cases > 0, Population > 500000) %>%
    filter(!is.na(fullp)) %>% 
    select(Country_Region, Province_State, cases, deaths,fullp) %>%
    ungroup() %>%
    slice_min(fullp, n=10)

top_10 <- state_cases %>%
    filter(date==ymd("2021/7/1"), cases > 0, Population > 500000) %>%
    filter(!is.na(fullp)) %>% 
    select(Country_Region, Province_State, cases, deaths,fullp) %>%
    ungroup() %>%
    slice_max(fullp, n=10)

bottom_10 

top_10 

```
On July 1st the lowest vaccinated states all had a rate below 40%, 
ranging from around 32.7% in Alabama to 38.8% in South Carolina.

The highest were all above 54%, ranging from 54.2% in New York 
to 65.2% in Rhode Island.

Also, on July 1st some of the states with high vaccinations rates already have 
a high number of total cases and deaths from earlier phases of the pandemic, 
for example New York and New Jersey. 

To account for this, I will use these numbers as the baseline and only look at 
new cases after that date.



## Analyse new case numbers in selected states

I will now calculate how many new cases these states have added after 7/1/21. 
To account for the different size of the states, I also calculate new 
cases/deaths per 100,000 residents.


```{r}
bottom_10 <- bottom_10 %>%
    rename (base_cases = "cases", base_deaths = "deaths",
            base_fullp = "fullp")

top_10 <- top_10 %>%
    rename (base_cases = "cases", base_deaths = "deaths",
            base_fullp = "fullp")

bottom_10_cases <- state_cases %>%
    filter(Province_State %in% bottom_10$Province_State, date>=ymd("2021/7/1")) %>%
    left_join(bottom_10,by = c("Province_State", "Country_Region")) %>%
    mutate(new_cases = cases - base_cases,
           new_deaths = deaths - base_deaths, 
           new_casesph = new_cases * 100000 / Population,
           new_deathsph = new_deaths * 100000 / Population,
           category = "Bottom 10")

top_10_cases <- state_cases %>%
    filter(Province_State %in% top_10$Province_State, date>=ymd("2021/7/1")) %>%
    left_join(top_10,by = c("Province_State", "Country_Region")) %>%
    mutate(new_cases = cases - base_cases,
           new_deaths = deaths - base_deaths, 
           new_casesph = new_cases * 100000 / Population,
           new_deathsph = new_deaths * 100000 / Population,
           category = "Top 10")
```


```{r}
    ggplot(bottom_10_cases, aes(x=date, y=new_casesph)) +
    geom_point(aes(color=Province_State)) +
    geom_line(aes(color=Province_State)) +
    ylim(0,3000) +
    labs(title = "Covid-Cases since 2021/7/1 in states with low vaccination rates", y="Number of new cases per 100K", x=NULL) 

    ggplot(top_10_cases, aes(x=date, y=new_casesph)) +
    geom_point(aes(color=Province_State)) +
    geom_line(aes(color=Province_State)) +
    ylim(0,3000) +
    labs(title = "Covid-Cases since 2021/7/1 in states with high vaccination rates", y="Number of new cases per 100K", x=NULL) 
  
    
```

The diagrams show that the states with the lowest vaccination rates as per 
July 1st have had consistently more new cases in the following months than the 
states with the highest vaccination rates. 

By the middle of August even Idaho, which has the lowest incidence rate of 
these 10 states has added more than 5000 new case per 100,000 residents. 
5 of theses states had more than 1000 new cases and Louisiana and Arkansas even 
reached 2000 new cases in the second half of August. 

In comparison, the case load in the highest vaccinated states overall looks 
much lower. 

In the 2nd half of August only Washington and New Mexico top 500 new infections
per 100,000 residents.

The graph also shows only moderate growth over the course of these two months.



# Modelling of COVID cases based on vaccinations 

I will now analyze if, as the graphs above seem to indicate, there is a 
relationship between the vaccination rate in a state and the number of 
new cases.

To build a model, I will first calculate for all states how many cases they 
have added since 7/1/2021.


```{r}
july1_cases <- state_cases %>%
    filter(date==ymd("2021/7/1"), cases > 0, Population > 500000) %>%
    filter(!is.na(fullp)) %>% 
    select(Country_Region, Province_State, cases, deaths, fullp) %>%
    rename (base_cases = "cases", base_deaths = "deaths",
            base_fullp = "fullp")

latest_cases <- state_cases %>%
    filter(!is.na(fullp)) %>% 
    filter(cases > 0, Population > 500000) %>%
    slice_max(date)

latest_cases <- latest_cases %>%
    left_join(july1_cases, by=c("Province_State", "Country_Region")) %>%
    mutate(new_casesph = (cases - base_cases) * 100000 / Population, 
           new_deathsph = (deaths - base_deaths) * 100000 / Population)

```


Which states have added the most and least cases since July?

```{r}
latest_cases %>%
    ungroup() %>%
slice_max(new_casesph, n=10 ) %>%
    select(Province_State, new_casesph, fullp )

latest_cases %>%
    ungroup() %>%
slice_min(new_casesph, n=10 ) %>%
    select(Province_State, new_casesph, fullp )

```

All of the 10 states with the highest number of cases added have a
vaccination rate below 50% as of 8/13/2021. 
5 of those have a rate lower than 40 %.

On the other hand 7 out of 10 states with the lowest number of cases have a 
vaccination rate above 50% and 5 of them have a rate above 55%.

I will plot a diagram to visualize the relationship of the number of new cases
and the vaccination rate.

```{r}
    ggplot(latest_cases, aes(x=fullp, y=new_casesph)) +
    geom_point(aes(color="New cases")) +
    geom_text(aes(label=ifelse(new_casesph>1200,Province_State,
                        ifelse(fullp>65,Province_State, "" ))),
              hjust=-0.1,vjust=-0.1, size=3) +
    xlim(35,75) +
    labs(title = "Number of Covid-Cases since 2021/7/1", x="Vaccination % ", 
         y="Total new cases per 100K")
```

It seems that states with a vaccination rate above 50% tend to have a 
lower rate of new cases. States with a rate lower than around 40% tend to have
a significantly higher number of new cases.

I will now try to model this behavior using a classification tree.

```{r}
mod = rpart(new_casesph ~ fullp, data=latest_cases) 
plot(mod, uniform=TRUE,margin=0.2)
text(mod, use.n=TRUE, cex=.8)

```

The model seems to confirm my previous finding that rates of 50% and 40% are an
indicator of whether the rate of new cases is trending significantly lower 
or higher respectively.

I will now use this model to calculate predictions of new number of cases based 
on the vaccination rate.


```{r}
latest_cases$pred <- predict(mod)


```

Now, I will plot the predictions and compare them with the actual number of 
new cases.


```{r}
    ggplot(latest_cases, aes(x=fullp, y=new_casesph)) +
    geom_point(aes(color="New cases")) +
    geom_line(aes(y=pred, color="New cases predicted")) +
    geom_text(aes(label=ifelse(new_casesph>1250,Province_State,
                        ifelse(fullp>65,Province_State, "" ))),
              hjust=-0.1,vjust=-0.1, size=3) +
    xlim(35,75) +
    labs(title = "Number of Covid-Cases since 2021/7/1", x="Vaccination % ", 
         y="Total new cases")

```

The diagram shows that the predicted number of cases somewhat reflects the 
number of cases. 

However, a lot of states with low vaccination rates (< 50 %) do not show the 
predicted number of cases. As shown above,  South Dakota, North Dakota and 
Michigan have among the 10th lowest number of cases, but their vaccination rates
are all below 50%.

On the other hand none of the states with the vaccination rate above 50% reaches more 
than 750 cases per 100,000 residents and most have even below 500 cases.
Florida, with a vaccination rate of 49.8%, seems to be the one big outlier.


# Bias Analysis

* Data Bias

    As the number of cases and vaccinations is reported by each state, 
    there might be different rules on how to test for and report COVID cases.
    Also, limited access to testing centers and medical facilities might lead to
    an under counting of cases.
    
    On the other hand, if vaccinated people get infected with the COVID-19 virus
    they might not show any symptoms and therefore will not get tested. 
    The numbers might therefore not correctly represent cases among the fully 
    vaccinated population.
    
    
* Personal Bias

    Given that I strongly believe in the efficacy of the COVID-19 vaccines, 
    my analysis might be biased in this regard.
    
    I might have overlooked other factors that could drive the new number 
    of cases lower, for example restrictions and adherence to CDC mask guidelines.
    
# Conclusions
1) States with a high rate of vaccinations (at least 50%) have been able to 
avoid a high increase in new cases during this new wave.   

2) Most of the states with a lower vaccination rate have seen a steep rise 
in the number of cases during the summer. 

    However, there is a high number of states with low vaccination rates 
that did not see a dramatic spike in cases.

    This indicates that other factors, maybe population density, adherence to CDC 
guidelines, etc. should be considered to build a more representative model.  

3) Even though it looks like the vaccination rate is not the only factor driving
the number of new infections, the analysis shows that a high vaccination rate 
is at least contributing to a lower number in cases. 

    Therefore, states and the federal government should look into measures that will 
encourage more people to get vaccinated. 



```{r}
sessionInfo()
```

