---
title: "NYC shootings"
date: "8/14/2021"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
library(tidyverse)
library(ggplot2)
```


# Statement of interest
For this project I will use the shooting incidence data provided by the city of 
New York to address the following questions:

1) How has the number of shooting incidents changed over the last years 
2) Does the number of shootings differ from Borough to Borough
3) Is the number of murders is related to the number of shooting incidents



# Import the data
I will import the data from the data repository of the City of New York.

```{r Import, echo=TRUE}
ny_shooting <- 
    read.csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
colnames(ny_shooting)

```


# Clean Up the data
I will extract the columns I need from the data set.

```{r Clean, echo=TRUE}
ny_shooting <- 
    select (ny_shooting, OCCUR_DATE, BORO, STATISTICAL_MURDER_FLAG) %>%
    rename(date = 'OCCUR_DATE', Boro = 'BORO', 
           Murder = 'STATISTICAL_MURDER_FLAG') %>%
    mutate(date = mdy(date))
summary(ny_shooting)
head(ny_shooting)
```

I have 15 years worth of data. To analyze further, I will count the cases per 
year and by borough.

Analyzing a time period of a year at a time, will also get rid of seasonal
fluctuations.

Also, I will translate the murder flag into 0,1 to be able to count the number
of murders.
```{r}

ny_shooting <- mutate(ny_shooting, Murder = ifelse(Murder == "true", 1, 0))
boro_year_count <- ny_shooting %>%
    group_by(Year = year(date), Boro) %>%
    summarize(Shootings=n(), Murders=sum(Murder), .groups = "drop")
    
head(boro_year_count)
```


# Visualize Data

To be able to compare the shootings in the different boroughs over time, 
I will plot them on a graph. 

To show the trend more clearly the smooth option is used.  

```{r}
ggplot(data=boro_year_count, aes(x=Year, y=Shootings, color=Boro)) +
    geom_point(aes(color= Boro)) +
    geom_smooth(se=FALSE) +
    scale_y_log10() +
    labs(title = "Shootings in the 5 Boroughs", y="Number of shootings", x=NULL)
    
    
```

Looking at the data a little closer a couple of things need to be further 
investigated.

1) The number of cases always seems to be highest in Brooklyn and lowest in 
Staten Island. 

    It might be helpful to look at the population of the different boroughs and 
look at shootings relative to the size of population.

2) The number of shootings has been almost consistently decreasing for the last
10 years. However, in 2020 there has been a significant uptick for each borough.

    Without additional data beyond that year, it is hard to deduce if this is an
anomaly maybe brought on by the COVID-19 epidemic or the beginning of a 
reversal of the positive trend of the previous years as the smoothed lines seem
to indicate.


## Import additional data

In order to calculate the shootings relative to the population of the borough,
I have found Census data published by the City of New York.

```{r }
boro_population <- 
read.csv("https://data.cityofnewyork.us/api/views/xywu-7bv9/rows.csv?accessType=DOWNLOAD")
colnames(boro_population)
```
As we only have shooting data spanning the range of 2006 to 2020, 

I will keep the population data from these years.


```{r}
boro_population <- select (boro_population, Borough, X2000, X2010, X2020) %>%
    rename(Boro="Borough")
head(boro_population)
```

A little clean-up, stripping leading blanks, capitalizing, to prepare data
before combining the two sets.

```{r}
boro_population$Boro = toupper(str_trim(boro_population$Boro))

boro_shooting <- merge(x=boro_year_count, y=boro_population, by= "Boro", 
                       all.x=TRUE, all.y=FALSE) 
head(boro_shooting)
```

Now, we can calculate the shootings, murders per 1 Mill residents. This will
allow us to better analyze if differences in the shootings are related to
the different Borough sizes.

Note, the census is only taken every 10 years. For years where no census data
is available, I will use the population data from the last census. 


```{r}
boro_shooting_corr <- mutate(boro_shooting, 
            ShootingsPer = ifelse(Year < year(ymd("2010-01-01")), 
            Shootings * 1000000 / X2000, ifelse(Year < year(ymd("2020-01-01")),
            Shootings * 1000000 / X2010, Shootings * 1000000 / X2020))) %>%
    mutate(boro_shooting,
            MurdersPer = ifelse(Year < year(ymd("2010-01-01")), 
            Murders * 1000000 / X2000, ifelse(Year < year(ymd("2020-01-01")),
            Murders * 1000000 / X2010, Murders * 1000000 / X2020)))
    
   
head(boro_shooting_corr)
```

## Visualize the adjusted Data

I will now plot the shootings per 1 Mill residents to see if there still is a 
big difference between the 5 boroughs.

```{r}
ggplot(data=boro_shooting_corr, aes(x=Year, y=ShootingsPer, color=Boro)) +
    geom_point(aes(color= Boro)) +
    geom_smooth(se=FALSE) +
    scale_y_log10() +
    labs(title = "Shootings in the 5 Boroughs", 
         y="Number of shootings per 1 Mio residents", x=NULL)
    
    
```

As we can see looking at the shootings relative to the size of the population,
Brooklyn actually does not have a significantly higher number of shootings than 
all of the other Boroughs. 

The Bronx has the highest number of shootings per 
residents. 

Brooklyn and the Bronx have more shootings per resident 
than the other three Boroughs. 

Manhattan, Queens and Staten Island are very 
similar in regards to the number of shootings per resident.


# Modelling of shooting and murder data 

Another interesting area to analyze is the outcome of the shootings. 
Naively, we would believe that an increased number of shootings will
lead to a higher numbers of murders.

I will plot the number of shootings and the number of murders to see if there 
is a possible correlation.


```{r}
boro_shooting_corr %>% ggplot() +
    geom_point(aes(x = ShootingsPer, y = MurdersPer), color = "blue") +
    labs(title = "NYC shootings and murders", 
    y="Number of murders per 1 Mio residents", 
    x="Shootings per 1 Mio residents")
```





The diagram shows that there is correlation between the number of shootings and deaths. 

I will build a linear model to see if it can be used to predict the number of murders given the number of shootings.


```{r}
linearMod <- lm( MurdersPer ~ ShootingsPer, data=boro_shooting_corr)  
summary(linearMod)
```
I will add a prediction for the number of murders based on the number of
shootings (both per Mill residents)

```{r}
boro_murder_pred <- boro_shooting_corr %>%
    mutate(predMurder = predict(linearMod))
view(boro_murder_pred)
```

Now, I will plot the predictions and compare them with the actual number of 
murders

```{r}
boro_murder_pred %>% ggplot() +
    geom_point(aes(x = ShootingsPer, y = MurdersPer), color = "blue") +
    geom_point(aes(x = ShootingsPer, y = predMurder), color = "red") +
    geom_text(aes(x = ShootingsPer, y = MurdersPer, 
            label=ifelse(MurdersPer > 100, paste(Boro,Year,sep= " "), "")),
               hjust=1.1,vjust=-0.1, size=3) +
    labs(title = "NYC shootings and murders",
    subtitle = "Actual vs. expected", 
    y="Number of murders per 1 Mio residents", 
    x="Shootings per 1 Mio residents")
```

It looks like my assumption of a proportional relationship between shootings
and murders holds true. 

The actual murder rate per 1 Mill residents (in blue) 
mostly follows the predicted number of murders (in red). 

However, one outlier
with over 100 murders should further be investigated.


# Bias Analysis

* Data Bias
    + Shooting Data
    
    It might be worth while looking at possible bias in the shooting data. 
    Do all the Boroughs have the same process in reporting the data? 
    + Population Data
    
    Given that census data relies on the participation of the citizens, it is 
    not a given that these numbers accurately reflect the number of residents.
    The Bronx and Brooklyn have a higher number of immigrants who are often not 
    accurately counted in Census data as they are reluctant to provide
    information to the government.

* Personal Bias

    Given that I personally lived in Brooklyn for a couple of years and have 
    witnessed a shooting myself during this time, I might be biased to believe 
    that Brooklyn is more dangerous than the other Boroughs. 
    
    This idea was confirmed after seeing the first graph showing Brooklyn 
    leading the number of shootings each year. 
    
    To mitigate this bias, I pulled in the population data in order to further 
    analyze the reasons for the high number of shootings in Brooklyn.
    
# Conclusions
1) Even adjusted for population size, the Bronx and Brooklyn have a higher 
number of shootings than Queens, Staten Island and Manhattan.

2) Even though the number of shootings has been on a downward trajectory for 
the last ten years, there has been a significant uptick in 2020. This holds 
true for all boroughs.

3) There is a direct relationship between the number of shootings and the 
number of murders. Most likely, measures put in place to reduce shootings
would also bring down the number of murders.
